# Longitudinal Analysis

```{r}
library(stringr); library(dplyr); library(GGally); library(corrplot); library(ggplot2); library(tidyr); library(MASS); library(boot); library(lme4)

RATS <- read.table("https://raw.githubusercontent.com/KimmoVehkalahti/MABS/master/Examples/data/rats.txt", header = TRUE, sep = '\t')
BPRS <- read.table("https://raw.githubusercontent.com/KimmoVehkalahti/MABS/master/Examples/data/BPRS.txt", sep  =" ", header = T)
```

### Part I: Analysis of RATS Data

This data was taken from a nutrition study conducted on three groups of rats. The groups were put on different diets, and each animal's body weight (grams) was recorded repeatedly (approximately) weekly, except in week 7 when two recordings were taken. This was conducted over a 9-week period.

```{r}
RATS$ID <- factor(RATS$ID)
RATS$Group <- factor(RATS$Group)
RATSL <- RATS %>%
  gather(key = WD, value = Weight, -ID, -Group) %>%
  mutate(Time = as.integer(substr(WD,3,4)))
str(RATSL)
```

```{r}
summary(RATSL)
```

### Graphical overview of data
```{r}
ggplot(RATSL, aes(x = Time, y = Weight, group = ID)) +
  geom_line(aes(linetype = Group)) +
  scale_x_continuous(name = "Time (days)", breaks = seq(0, 60, 10)) +
  scale_y_continuous(name = "Weight (grams)") +
  theme(legend.position = "top")
```

While there is variability within the groups, rats in group 1 showed slow weight gain, whereas rats in groups 2 and 3 started with a higher weight and also gained more weight over time. 

### Standardise the variables

```{r}
RATSL <- RATSL %>%
  group_by(Group) %>%
  mutate(stdweight = (Weight - mean(Weight))/sd(Weight) ) %>%
  ungroup()

summary(RATSL)
```

```{r}
ggplot(RATSL, aes(x = Time, y = stdweight, group = ID)) +
  geom_line(aes(linetype = Group)) +
  scale_x_continuous(name = "Time (days)", breaks = seq(0, 60, 10)) +
  scale_y_continuous(name = "Standardized Weight (grams)") +
  theme(legend.position = "top")
```

Once standardised we can see that the groups are on similar trajecties but maintain significant variance between the rats.

```{r}

n <- RATSL$Time %>% unique() %>% length()

# Summary data with mean and standard error  
RATSS <- RATSL %>%
  group_by(Group, Time) %>%
  summarise( mean = mean(Weight), se = sd(Weight)/sqrt(n) ) %>%
  ungroup()

# Glimpse the data
glimpse(RATSS)
```

```{r}
# Plot the mean profiles
ggplot(RATSS, aes(x = Time, y = mean, linetype = Group, shape = Group)) +
  geom_line() +
  scale_linetype_manual(values = c(1,2,3)) +
  geom_point(size=3) +
  scale_shape_manual(values = c(1,2,3)) +
  geom_errorbar(aes(ymin = mean - se, ymax = mean + se, linetype="1"), width=0.3) +
  theme(legend.position = c(0.8,0.8)) +
  scale_y_continuous(name = "mean(Weight) +/- se(Weight)")
```

Again, we can see that there is a large difference between the mean of group 1 and groups 2 and 3. Groups 2 and 3 were larger to begin with and are gaining weight faster.

```{r}
# Create summary data 
RATSL8S <- RATSL %>%
  group_by(Group, ID) %>%
  summarise( mean=mean(Weight) ) %>%
  ungroup()

# Glimpse the data
glimpse(RATSL8S)

# Draw a boxplot of the mean versus group
ggplot(RATSL8S, aes(x = Group, y = mean)) +
  geom_boxplot() +
  stat_summary(fun.y = "mean", geom = "point", shape=23, size=4, fill = "white") +
  scale_y_continuous(name = "mean(Weight), time")
```

This graph shows that group 1 had the lowest weight, group 3 had the highest weight and group 2 had the most variance in weight.

```{r}
# Add the baseline from the original data as a new variable to the summary data
RATSL8S2 <- RATSL8S %>%
  mutate(baseline = RATS$WD1)

# Fit the linear model with the mean as the response 
fit <- lm(mean ~ Group, data = RATSL8S2)

# Compute the analysis of variance table for the fitted model with anova()
anova(fit)
```

### Part II: Analysis of BPRS Data
This data consists of 40 male subjects who were randomly assigned to one of two treatment groups where each subject was rated on the brief psychiatric rating scale (BPRS) which is used to evaluate patients suspected of having schizophrenia. The BPRS was measured before treatment began on week 0 and then at weekly intervals until week 8. 

BPRS assesses the level of 18 symptom constructs including hostility, suspiciousness, hallucinations and grandiosity. Each symptom construct is rated from 1 (not present) to 7 (extremely severe).

```{r}
BPRS$treatment <- factor(BPRS$treatment)
BPRS$subject <- factor(BPRS$subject)
BPRSL <-  BPRS %>% gather(key = weeks, value = bprs, -treatment, -subject)
BPRSL <-  BPRSL %>% mutate(week = as.integer(substr(weeks,5,5)))
str(BPRSL)
```

```{r}
summary(BPRSL)
```

```{r}
ggplot(BPRSL, aes(x = week , y = bprs, linetype = subject)) +
  geom_line() +
  scale_linetype_manual(values = rep(1:10, times=4)) +
  facet_grid(. ~ treatment, labeller = label_both) +
  theme(legend.position = "none") + 
  scale_y_continuous(limits = c(min(BPRSL$bprs), max(BPRSL$bprs)))
```

First we plotted the data to identify the observations in each group while ignoring the longitudinal nature of the data.

The above plots show the different bprs scores gathered over an 8 week period for the different treatment groups.

### Create regression model
```{r}
# create a regression model 
BPRS_reg <- lm(bprs ~ week + treatment, data = BPRSL)

# print out a summary of the model
summary(BPRS_reg)
```
Next, we created a linear regression model with bprs score as the intercept and week and treatment as explanatory variables.

The regression analysis shows that there is a strong correlation between the intercept (bprs) and time (week) in treatment.

```{r}
# Create a random intercept model
BPRS_ref <- lmer(bprs ~ week + treatment + (1 | subject), data = BPRSL, REML = FALSE)

# Print the summary of the model
summary(BPRS_ref)
```
A random intercept is fit so that the linear regression for each bprs score can differ in intercept for other subjects, thereby accounting for individual differences in the subject profiles.

```{r}
# create a random intercept and random slope model
BPRS_ref1 <- lmer(bprs ~ week + treatment + (week | subject), data = BPRSL, REML = FALSE)

# print a summary of the model
summary(BPRS_ref1)

# perform an ANOVA test on the two models
anova(BPRS_ref1, BPRS_ref)
```

```{r}
BPRS_ref2 <- lmer(bprs ~ week * treatment + (week | subject), data = BPRSL, REML = FALSE)

# print a summary of the model
summary(BPRS_ref2)

# perform an ANOVA test on the two models
anova(BPRS_ref2, BPRS_ref1)

# Create a vector of the fitted values
Fitted <- fitted(BPRS_ref2)

# Create a new column fitted to BPRSL
BPRSL <- BPRSL %>%
  mutate(Fitted)

# draw a plot
ggplot(BPRSL, aes(x = week, y = Fitted, group = subject)) +
  geom_smooth(aes(linetype = subject)) +
  scale_x_continuous(name = "Time",) +
  scale_y_continuous(name = "Fitted bprs") +
  theme(legend.position = "top")

```

This chart shows that the longer a subject is in treatment, the lower their bprs score. We can assume therefore that treatment is effective at reducing the presentation of symptoms which may indicate schizophrenia.