# Classification and clustering

```{r}
# Load libraries
library(MASS); library(tidyr); library(dplyr); library(ggplot2); library(GGally); library(corrplot)
# Load data
data("Boston")

```

*ABOUT THE DATA HERE- Explore the structure and the dimensions of the data and describe the dataset briefly, assuming the reader has no previous knowledge of it.*

```{r}
str(Boston)
summary(Boston)

```

```{r}
pairs(Boston)
cor_matrix<-cor(Boston)
print(cor_matrix)
corrplot(cor_matrix, method="circle", type = "upper", cl.pos = "b", tl.pos = "d", tl.cex = 0.6) %>% round(digits = 2)
```

*ABOUT THE DATA HERE - show summaries of the variables in the data. Describe and interpret the outputs, commenting on the distributions of the variables and the relationships between them.*

### Standardize dataset

```{r}
boston_scaled <- scale(Boston)
summary(boston_scaled)
class(boston_scaled)
boston_scaled <- as.data.frame(boston_scaled)
```

*WRITE ME - How did the variables change?*

### Divide the dataset into train and test sets
```{r}
summary(boston_scaled$crim)
bins <- quantile(boston_scaled$crim)
bins
crime <- cut(boston_scaled$crim, breaks = bins, include.lowest = TRUE, label = c("low", "med_low", "med_high", "high"))
table(crime)
boston_scaled <- dplyr::select(boston_scaled, -crim)
boston_scaled <- data.frame(boston_scaled, crime)
n <- nrow(boston_scaled)
ind <- sample(n,  size = n * 0.8)
train <- boston_scaled[ind,]
test <- boston_scaled[-ind,]
```

### Linear discriminant analysis
```{r}
lda.fit <- lda(crime ~ ., data = train)
lda.fit
lda.arrows <- function(x, myscale = 1, arrow_heads = 0.1, color = "red", tex = 0.75, choices = c(1,2)){
  heads <- coef(x)
  arrows(x0 = 0, y0 = 0, 
         x1 = myscale * heads[,choices[1]], 
         y1 = myscale * heads[,choices[2]], col=color, length = arrow_heads)
  text(myscale * heads[,choices], labels = row.names(heads), 
       cex = tex, col=color, pos=3)
}
classes <- as.numeric(train$crime)
plot(lda.fit, col = classes, pch = classes, dimen = 2)
lda.arrows(lda.fit, myscale = 1)
```

```{r}
correct_classes <- test$crime
test <- dplyr::select(test, -crime)
```


