# Regression and model validation

```{r}
# Load libraries
library(ggplot2)
library(GGally)

# Import data
lrn14 <- read.table("http://s3.amazonaws.com/assets.datacamp.com/production/course_2218/datasets/learning2014.txt", sep=",", header=TRUE)
```

The data used in this analysis is from an international survey of Approaches to Learning collected from 3.12.2014 - 10.1.2015. Survey was originally done in Finnish and then translated to English. Data was used to explore the relationship between learning approaches and student achievements. 

A description of the variables in the dataset includes: gender (male or female), age (in years, derived from date of birth), attitude (global attitude toward statistics), deep (deep approach to learning), stra (strategic approach to learning), surf (surface approach to learning) and points (exam points). The data has 166 observations and 7 variables as well as 166 rows and 7 columns.

```{r}
str(lrn14) 
dim(lrn14)
```

```{r}
# Scatter plot of matrix with variables, excluding gender
pairs(lrn14[-1])
```

```{r}
# Summaries of the variables
summary(lrn14)
```

```{r}
# Plot matrix of variables
ggpairs(lrn14, lower = list(combo = wrap("facethist", bins = 20)))
```

#### Analysis of variables
An analysis of the variables shows that attitude, deep, stra, and surf have relatively normal distribution. The variables with the highest correlations to the output variable (points) are attitude (0.437), stra (0.146) and surf (-0.144). There is also a correlation between deep and surf (-0.324). Finally, it's interesting to note that there are twice as many women in this dataset as men.


### First regression model
This model looks at the correlation between the predictor variables attitude, stra and surf and the outcome variable points. The predictor variables were chosen based on their relatively high correlation values when compared to points.

```{r}
# Create a regression model with multiple explanatory variables
model1 <- lm(points ~ attitude + stra + surf, data = lrn14)

# Summary of the model
summary(model1)
```

#### Analysis of first model
This model explains .1927 of the variance present. Variables stra and surf both have p-levels which exceed 0.05 and are therefore not statistically significant. Attitude does correlate with points, however, and will therefore be included in the second model.

### Second regression model
Because stra and surf had no statistically significant correlation with points they were excluded from the second model. This model looks at the correlation between points and attitude.
```{r}
# Create a second regression model with statistically relevant variables
model2 <- lm(points ~ attitude, data = lrn14)

qplot(attitude, points, data = lrn14) + geom_smooth(method = "lm")

# Summary of the model
summary(model2)
```

#### Analysis of second model
This model shows a positive correlation between points and attitude - as the attitude towards statistics increases so do the number of exam points.

R-squared measures the closeness of data to the fitted regression line. In this case, the multiple R-squared is 0.1906, which is slightly lower than in the first model.

#### Diagnostic plots
These plots are used to explore the validity of the model assumptions.

```{r}
par(mfrow = c(2,2))
plot(model2, which = c(1:2,5))
```

**Residuals vs Fitted - **
The Residuals vs Fitted plot is based on the assumption that the size of error doesn't depend on explanatory variables. The plot shows whether there is a problem with this assumption. There is no pattern in the data, therefore the model adheres to the assumption.


**Q-Q plot - **
Q-Q plots are used to prove whether a model adheres to the normality of errors assumption whereby variables are normally distributed. The variables in this model fit relatively well to the line and therefore does not violate the normality assumption.


**Residuals vs Leverage - **
Leverage of observations measures how much impact a single observation has on the model. The Residuals vs Levarage plot for this model indicates regular leverage.
