# Dimensionality Reduction

```{r}
# Load libraries
library(stringr); library(dplyr); library(GGally); library(corrplot); library(ggplot2); library(tidyr)

# Load data
human <- read.table("http://s3.amazonaws.com/assets.datacamp.com/production/course_2218/datasets/human2.txt", sep=",", header=TRUE)

```
### Description of data
Data is taken from the United Nations Development Programme. It combines the following two indices: 

Human Development Index (HDI): http://hdr.undp.org/en/composite/HDI

Gender Inequality Index (GNI): http://hdr.undp.org/en/composite/GII

For this analysis the two data sets have been combined by the variable 'country' while variables unnecessary for this analysis have been removed. Additionally, the variable capturing Gross National Income (GNI) has been converted to a numerica variable.

Definitions of the variables are as follows:

GNI - Gross National Income per capita

Life.Exp - Life expectancy at birth

Edu.Exp - Expected years of schooling 

Mat.Mor - Maternal mortality ratio

Ado.Birth - Adolescent birth rate

Parli.F - Percetange of female representatives in parliament

Edu2.FM - Proportion of females with at least secondary education / Proportion of males with at least secondary education

Labo.FM - Proportion of females in the labour force / Proportion of males in the labour force

### Graphical overview of data
```{r}
summary(human)

# visualize human variables
ggpairs(human)

# compute and visualise the correlation matrix
cor_matrix <- cor(human) %>% round(digits=2)
cor_matrix

corrplot(cor_matrix, method="circle", type="upper", cl.pos = "b", tl.pos = "d",tl.cex = 0.6)

```

For the most part the variables are skewed, with the exception of ratio of expected secondary education for men and woman and expected years of schooling. There are some outliars in the data as seen in the scatterplots. Life expectancy is between 49 and 83 years of age. 

Strong positive correlations are seen between life expectancy and expected years of schooling (0.79) and maternal mortality and adolescent birth rate (0.76). On the other hand, strong negative correlations were seen between life expectancy and maternal mortality (-0.86), expected education and maternal mortality (-0.74) and life expectancy and adolescent birth rate (-0.73).

### Perform PCA on non-standardized data
```{r}
# perform principal component analysis (with the SVD method)
pca_human <- prcomp(human)

# draw a biplot of the principal component representation and original variables
biplot(pca_human, choices = 1:2, cex = c(0.8, 1), col = c("grey40", "deeppink2"))
```

### Perform PCA on standardized data
```{r}
# standardize the variables
human_std <- scale(human)

# perform principal component analysis (with the SVD method)
pca_human <- prcomp(human_std)

# draw a biplot of the principal component representation and the original variables
biplot(pca_human, choices = 1:2, cex = c(0.8, 1), col = c("grey40", "deeppink2"))
```

The biplot looks significantly different between the unstandardized and standardized data. The biplot of the unstandardized data sees the results bunched at the top end of the plot with GNI as an outlier as it is on a significantly different scale. Once standardized, we can begin to interpret the data. 

There appear to be three principle components within the standardized data which are as follows:

1) There is a relationship between maternal mortality and adolescent birth rates. This is intuitive as very young mothers may suffer more from complications in child birth.

2) Another relationship is between the ratio between men and women in the work force and the percentage of female representatives in parliament. This is also intuitive as the more women there are in the work force the more likely it is that women will run for elected office.

3) The final relationship presents a more complex relationship between expected level of education, ratio of men and women in secondary education and life expectancy.

### Multiple Correspondence Analysis
```{r}
library(FactoMineR)
data(tea)
```

```{r}
keep_columns <- c("Tea", "How", "how", "sugar", "where", "lunch")

# select the 'keep_columns' to create a new dataset
tea_time <- dplyr::select(tea, one_of(keep_columns))

# look at the summaries and structure of the data
str(tea_time)
summary(tea_time)

# visualize the dataset
gather(tea_time) %>% ggplot(aes(value)) + geom_bar() + theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 8)) + facet_wrap("key", scales = "free") 
```

### Run a multiple correspondance analysis on tea data

```{r}
# multiple correspondence analysis
mca <- MCA(tea_time, graph = FALSE)

# summary of the model
summary(mca)

# visualize MCA
plot(mca, invisible=c("ind"), habillage = "quali")
```

Based on the plot we can see a relationship between those who buy unpackaged tea in tea shops (and possibly prefer green tea) as well as those who buy tea bags in chain stores. We can also see that both Dim.1 and Dim.2 have something to do with how people drink tea (teabag, loose or both) and where they purchase the tea, whereas Dim.3 describes the type of tea, how they prepare it and amount of sugar.